Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-11-04 fstats_pca_lewontin.tex

\section{Discussion}
Particularly for the analysis of human genetic variation with a large number of individuals with heterogeneous relationships, $F$-statistics  are  a powerful tool to describe population genetic diversity. Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA-plot. This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots. As PCA is often ran in an early step in data analysis, this may aid in generation of hypotheses that can be more directly evaluated using generative models, typically using a lower number of populations. It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots. In many cases, differences between the two data summaries will be due to variation on higher PCs. In this case, plotting additional PCs, or further subsetting the data to a more local set of populations seems prudent.

The other cause for disagreements between $F$-statistics and PCA are differences in assumptions. The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized. In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies. In contrast, most data analyses have to grapple with missing data, SNP are often weighted according to their allele frequency and observed, individual-level genotypes are used as the basis of PCA.

The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}. Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}. In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation. These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}, and a PCA-based approach to estimating $F$-statistics may provide higher accuracy than naive approaches because it allows using data from related populations.

The normalization of SNPs is similarly a matter of convention. The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable; and the same would hold if SNPs were weighted. The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}. As genetic differentiation between human populations is low, the  normalization used may matter little in practice \citep{felsenstein1973}, but using consistent normalization between $F$-statistics and PCA-based analysis seems advisable since it removes normalization as a source of errors. .

The third issue of individual-based vs population-based analysis is similarly a matter of interpretation. For $n$ samples, the number of possible $F$-statistics is on the order of $n^4$, and so the number of statistics is kept low by grouping individuals into populations. This is, however, not necessary, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}. Here, individual-based PCA can provide some guidance for grouping samples: Since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on a PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.

The final difference between the $F$-statistic-based PCA used here and individual-based PCA is on the usage of estimated allele frequencies versus individual-based genotypes. The fact that PCA does not distinguish between sample-based errors and the underlying structure is a well-known drawback of standard PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise. Probabilistic PCA is one class of approaches that aim to separate the population structure from individual-level noise \citep{agrawal2020}, and it seems likely that probabilistic PCA would yield a representation of the data that corresponds more closely aligned with $F$-statistics than regular PCA.

Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions that are partially arbitrary, and partially  explained by the focus of PCA on exploratory data analysis. Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if PCA and $F$-statistics are used in a way such that the results are comparable to each other.

Previous interpretation of PCA in the context of population genetic models have focused on simple models such as trees \citep{cavalli-sforza1975}, homogeneous spatial models \citep{novembre2008a} and discrete-population models \citep{franccois2021}. My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs. One consequence is that the results here are not impacted by sample ascertainment, sample sizes or number of principal components analyzed, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis. From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.

As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}. However, in many regions, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}. This provides a challenge for interpretation; as many $F_3$ and $F_4$ statistics may indicate gene flow. In my example (Figure \ref{fig:f3}A), most Southern European populations are ``admixed'' between Basques and Turkish, but a more accurate model might be one of continuous variation where Basque and Turkish lie on one of multiple gradients; which is more directly visualized with PCA. There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. One example of this is the world foci data set used here, which contains only 33 populations from across the world, and which is well-approximated by two PCs. However, this ascertainment misses a large amount of variation; a more dense sampling would show that in many places human genetic diversity is very gradual and multi-layered \citep{lazaridis2014, peter2020a}. The PCA-based  interpretation offers an alternative that trades interpretability for robustness. Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow. Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.

>>>>

\section{Discussion}
Particularly for the analysis of human genetic variation with a large number of individuals with heterogeneous relationships, $F$-statistics  are  a powerful tool to describe population genetic diversity. Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA-plot.

\subsection{more PCs}
This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots. As PCA is often ran in an early step in data analysis, this may aid in generation of hypotheses that can be more directly evaluated using generative models, typically using a lower number of populations. It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots. In many cases, differences between the two data summaries will be due to variation on higher PCs. In this case, plotting additional PCs, or further subsetting the data to a more local set of populations seems prudent.

\subsection{Assumptions}
The other cause for disagreements between $F$-statistics and PCA are differences in assumptions. The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized. In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies. In contrast, most data analyses have to grapple with missing data, SNP are often weighted according to their allele frequency and observed, individual-level genotypes are used as the basis of PCA.

\subsubsection{missing data}
The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}. Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}. In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation. These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}, and a PCA-based approach to estimating $F$-statistics may provide higher accuracy than naive approaches because it allows using data from related populations.

\subsubsection{normalization}
The normalization of SNPs is similarly a matter of convention. The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable; and the same would hold if SNPs were weighted. The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}. As genetic differentiation between human populations is low, the  normalization used may matter little in practice, but should be explored in future work \citep{felsenstein1973}.

\subsubsection{estimated vs. observed allele frequencies}
The third difference between $F$-statistics and standard  PCA is on the usage of estimated allele frequencies versus individual-based genotypes. The fact that PCA does not distinguish between sample-based errors and the underlying structure is a well-known drawback of standard PCA, and applying the theory presented here to individual-based PCA would result in $F$-statistics that incorporate some sampling noise. Probabilistic PCA is one class of approaches that aim to separate the population structure from sampling noise \citep[e.g.][]{agrawal2020}. It seems likely that probabilistic PCA would yield a representation of the data that corresponds more closely aligned with $F$-statistics than regular PCA.

\subsubsection{individual vs. population-based analyses}
The final issue is that PCA is commonly run on individual-based data, whereas $F$-statistics often group individuals into populations. This is not necessary; population-based PCA has been the default in the past \citep{cavalli-sforza1994}, and $F$-statistics are often applied to individuals \citep[e.g.][]{green2010, massilani2020, yang2020}. Thus, more commonly the choice between individual-based and population-based approaches is in the goal of an analysis. Grouping individuals into populations introduces a major assumption, that can e.g. be justified using an individual-based PCA. In particular, since $F$-statistics assume individuals are randomly drawn from a population, they should form tight clusters on an individual-based PCA-plot, otherwise population substructure becomes a possible alternative model for negative $F_3$-statistics and non-zero $F_4$-statistics \citep{peter2016}.

\subsubsection{summary}
Thus, while the version of PCA used here differs from that proposed by \cite{patterson2012}, the differences are largely due to conventions, or due to different goals for a particular analysis. Particularly for studies where the description of population structure is a major focus, results might be easier to interpret if  conventions regarding missing data, normalization and estimation of allele frequencies are used consistently between $F$-statistics and PCA.

\subsection{Previous Interpretations}
Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2021}. My interpretation here is different in that the utility of PCA is to simplify the geometry of the data. One consequence is that the results here are not impacted by sample ascertainment, sample sizes or the number of principal components analyzed, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis. From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.

The two data sets I analyzed here suggest that two PCS (for the World data set) respectively three PCS (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:f3}E, \ref{fig:f4}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure. On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:f3}BC), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs. Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs.

\subsection{broader scope}

As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}. However, in many parts of the world, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}. This provides a challenge for interpretation; as most $F_3$ and $F_4$-statistics will indicate departures from treeness. A naive interpretation of the $F$-statistics from my Eurasian example  (Figure \ref{fig:f3}A) would identify a substantial fraction of Europeans as (significantly) admixed between Finnish and Sardinians. In contrast, PCA reveals that the variation in this data set is not due to a single event, and so an arguably better description of the data set is one where Finnish and Sardinians lie on opposite ends of a more gradually structured population.

This trade-off between $F$-statistics, which allow for significance teste tension between

There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. One example of this is the world foci data set used here, which contains only 33 populations from across the world, and which is well-approximated by two PCs. However, this ascertainment misses a large amount of variation; a more dense sampling would show that in many places human genetic diversity is very gradual and multi-layered \citep{lazaridis2014, peter2020a}. The PCA-based  interpretation offers an alternative that trades interpretability for robustness. Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow. Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.
